---
title: "Homework 02 - Data Visualization, Wrangling and Advanced Regression"
author: "Sophie Xu"
date: "February 16, 2025"
output: 
  html_document: 
    highlight: kate
    theme: cosmo
---

```{r setup, message=FALSE, echo=FALSE, warning=FALSE}
# Scraping
library(rvest)
library(xml2)
library(base)
library(stringr)

# Dataframe
library(data.table)
library(dtplyr)
library(dplyr)

# Visualization
#install.packages("plotly")
library(leaflet)
library(tidyverse)
library(ggplot2)
library(gridExtra)
library(viridis)
library(plotly)

# Table
knitr::opts_chunk$set(include  = TRUE)
library(kableExtra)

# Model
library(mgcv)
```

### Introduction

This assignment examines the relationship between alcohol consumption and life expectancy through data analysis. The primary questions of interest are: 
* Is there an association between life expectancy and alcohol consumption?
* Is there a difference in the association between life expectancy and alcohol for males and females?
* How have life expectancy and alcohol consumption changed over time?
To answer these questions, we perform data wrangling, visualization, and regression analysis.

### 1 Data Wrangling

For both datasets, we renamed variables for clarity and extracted relevant variables: country, country code, population, year, sex, life expectancy, and alcohol consumption. In the life expectancy dataset, we consolidated male and female life expectancy into a single column with a corresponding sex variable. The data spans from 1950 to 2023, though some population and life expectancy values are missing. For the alcohol consumption dataset, we excluded records labeled “both sexes” for simplicity. This dataset spans from 2000 to 2019 and has no missing values. We also categorized alcohol consumption into levels based on quartiles. After merging the datasets by country, sex, and year, we obtained a final dataset with 7,492 observations.

```{r, eval=TRUE, echo=FALSE, message=FALSE}
# Load Data
df_life <- read.csv("https://ourworldindata.org/grapher/life-expectancy-of-women-vs-life-expectancy-of-men.csv?v=1&csvType=full&useColumnShortNames=true")
df_alcohol <- data.table::fread("WHOAlcoholTotalPerCapita_2021-09-20v2.csv")
```

```{r, eval=TRUE, echo=FALSE, message=FALSE}
# Data Cleaning: Life Expectancy
life <- df_life |>
  select(-c(Entity, owid_region))

life <- life |>
  rename("Population" = "population__sex_all__age_all__variant_estimates")

life <- life |>
  pivot_longer(
    cols = starts_with("life_expectancy__sex_"),
    names_to = "Sex",
    values_to = "Life Expectancy"
  )

life <- life |>
  mutate(Sex = case_when(
    grepl("life_expectancy__sex_female__age_0__variant_estimates", Sex) ~ "Female",
    grepl("life_expectancy__sex_male__age_0__variant_estimates", Sex) ~ "Male"
  ))
```

```{r, eval=FALSE, echo=FALSE, message=FALSE}
summary(life)
head(life)
```

```{r, eval=TRUE, echo=FALSE, message=FALSE}
# Data Cleaning: Alcohol Consumption
alcohol <- df_alcohol |>
  select(-c("WHO Region Code", "WHO Region", "Alcohol total per capita (15+) consumption in liters (string)"))

alcohol <- alcohol |>
  filter(Sex != "Both sexes")

alcohol <- alcohol |>
  rename("Code" = "Country Code",
         "Consumption" = "Alcohol total per capita (15+) consumption in liters (numeric)",
         "Lower Bound" = "Alcohol total per capita (15+) consumption in liters (low estimation)",
         "Upper Bound" = "Alcohol total per capita (15+) consumption in liters (high estimation)"
         )

alcohol <- alcohol |> select(-c("Lower Bound", "Upper Bound")) # seems like we don't need it
```

```{r, eval=FALSE, echo=FALSE, message=FALSE}
summary(alcohol)
head(alcohol)
```

```{r, eval=TRUE, echo=FALSE, message=FALSE}
# Merge Data
df <- merge(life, alcohol, by = c("Code", "Year", "Sex")) # all = TRUE
```

```{r, eval=FALSE, echo=FALSE, message=FALSE}
summary(df)
head(df)
```

```{r, eval=TRUE, echo=FALSE, message=FALSE}
# Report Cleaning Results
knitr::kable(summary(df))

table_data <- data.frame(
  Stage = c("Life Expectancy: Before Merge", "Life Expectancy: After Collapsing Sex", 
            "Alcohol: Before Filter", "Alcohol: After filter", "Data: After Merge"),
  Observations = c(
    dim(df_life)[1],
    dim(life)[1],
    dim(df_alcohol)[1],
    dim(alcohol)[1],
    dim(df)[1]
  )
)
knitr::kable(table_data, col.names = c("Stage", "Number of Observations"))
```

To better understand the data, we calculated the mean and standard deviation of life expectancy and alcohol consumption by year and sex. Life expectancy has increased over time, with females consistently living longer than males (e.g., ~75 vs. ~70 years in 2019). The standard deviation of life expectancy has decreased, with the gender gap narrowing to approximately eight years by 2019. Alcohol consumption, in contrast, has remained relatively stable, with females consistently consuming less than males (~2.5 vs. ~9.3 liters per capita in 2019). Interestingly, the standard deviation of alcohol consumption is nearly equal to the mean, though it declined slightly from 2000 to 2019.

```{r, eval=TRUE, echo=FALSE, message=FALSE}
# Report Grouped Mean and SD
summary_table <- df |>
  group_by(Year, Sex) |>
  summarize(
    Mean_Life_Expectancy = mean(`Life Expectancy`, na.rm = TRUE),
    SD_Life_Expectancy = sd(`Life Expectancy`, na.rm = TRUE),
    Mean_Alcohol_Consumption = mean(Consumption, na.rm = TRUE),
    SD_Alcohol_Consumption = sd(Consumption, na.rm = TRUE)
  ) |>
  ungroup()
knitr::kable(summary_table, 
             col.names = c("Year", "Sex", 
                           "Mean Life Expectancy", "SD Life Expectancy", 
                           "Mean Alcohol Consumption", "SD Alcohol Consumption"),
             digits = 3)
```

```{r, eval=TRUE, echo=FALSE, message=FALSE}
# Compute Quartiles
consumption_quartiles <- df |>
  group_by(Sex) |>
  summarize(
    Q1 = quantile(Consumption, 0.25, na.rm = TRUE),
    Q3 = quantile(Consumption, 0.75, na.rm = TRUE)
  )

# Create Categorical Variable
df <- df |>
  left_join(consumption_quartiles, by = "Sex") |>
  mutate(
    `Consumption Level` = case_when(
      Consumption <= Q1 ~ "Low",
      Consumption > Q1 & Consumption <= Q3 ~ "Medium",
      Consumption > Q3 ~ "High",
      TRUE ~ NA_character_
    ),
    `Consumption Level` = factor(`Consumption Level`, levels = c("Low", "Medium", "High"))
  )
df <- df |> select(-c("Q1", "Q3"))

# Validate by Summary Table
validate <- df |>
  group_by(Sex, `Consumption Level`) |>
  summarize(
    Min_Consumption = min(Consumption, na.rm = TRUE),
    Max_Consumption = max(Consumption, na.rm = TRUE),
    Num_Observations = n()
  ) |>
  ungroup()
knitr::kable(validate, 
             col.names = c("Sex", "Consumption Level", "Min Consumption", 
                           "Max Consumption", "Number of Observations"))
```

### 2 Exploring the Data

We first examined variable types and converted character variables to factors for ease of use. We confirmed that each country has a unique country code but found missing data for two countries between 2000 and 2006, which we chose to retain. A summary of the dataset revealed a balanced distribution of sexes, a wide range of life expectancy (12.38 to 87.43 years), and most observations falling into the medium alcohol consumption category. Exploratory visualizations supported these findings. A histogram of life expectancy by alcohol consumption level indicated that life expectancy modes were similar across all three consumption levels, suggesting that alcohol consumption may not be a strong predictor of life expectancy variation.

```{r, eval=FALSE, echo=FALSE, message=FALSE}
## check dimensions & variable types
str(df) 
```

```{r, eval=TRUE, echo=FALSE, message=FALSE}
## check dimensions & variable types
df <- df |> 
  mutate(across(where(is.character), as.factor))
```

```{r, eval=TRUE, echo=FALSE, message=FALSE}
## take a closer look & validate
check_mapping <- paste("Unique Codes:", length(unique(df$Code)), 
            "Unique Countries", length(unique(df$Country)))
#print(check_mapping)
country_counts <- df |> group_by(Code) |> summarise(Count = n())
#summary(country_counts)
missing <- country_counts[country_counts$Count != 40, ]
#summary(df[df$Code %in% missing$Code, ])
#summary(life[life$Code %in% missing$Code, ])
m <- alcohol[alcohol$Code %in% missing$Code, ]
ggplot(m, aes(x = Year, y = Consumption, color=Sex, shape = Code)) +
  geom_point(alpha = 0.5) +
  labs(x = "Year", y = "Alcohol Consumption", title = "Alcohol Consumption Over Time") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, size = 11, face = "bold"))
```

```{r, eval=TRUE, echo=FALSE, message=FALSE}
# summary statistics
knitr::kable(summary(df))
```

```{r, eval=TRUE, echo=FALSE, message=FALSE}
# visualizations
ggplot(df, aes(x = Year, y = `Life Expectancy`, color=Sex)) +
  geom_point(alpha = 0.5) +
  labs(x = "Year", y = "Life Expectancy", title = "Life Expectancy Over Time") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, size = 11, face = "bold"))
```

```{r, eval=TRUE, echo=FALSE, message=FALSE}
ggplot(df, aes(x = Consumption, fill = Sex)) +
  geom_histogram(bins = 20, alpha = 0.6, color = "white", position = "identity") +
  
  geom_vline(data = consumption_quartiles |> filter(Sex == "Female"), 
             aes(xintercept = Q1), linetype = "dashed", color = "red", alpha = 0.3) +
  geom_vline(data = consumption_quartiles |> filter(Sex == "Female"), 
             aes(xintercept = Q3), linetype = "dashed", color = "red", alpha = 0.3) +
  geom_vline(data = consumption_quartiles |> filter(Sex == "Male"), 
             aes(xintercept = Q1), linetype = "dashed", color = "blue", alpha = 0.3) +
  geom_vline(data = consumption_quartiles |> filter(Sex == "Male"),
             aes(xintercept = Q3), linetype = "dashed", color = "blue", alpha = 0.3) +

  geom_rect(data = consumption_quartiles, 
            aes(xmin = Q1, xmax = Q3, ymin = 0, ymax = Inf, fill = Sex), 
            alpha = 0.2, inherit.aes = FALSE) +
  
  labs(x = "Alcohol Consumption", y = "Frequency", 
       title = "Distribution of Alcohol Consumption by Sex") +
  scale_fill_manual(values = c("Female" = "pink", "Male" = "lightblue")) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, size = 11, face = "bold"))
```
```{r, eval=TRUE, echo=FALSE, message=FALSE}
ggplot(df, aes(x = `Life Expectancy`, fill = `Consumption Level`)) +
  geom_histogram(bins = 30, alpha = 0.8, color = "white") +
  labs(x = "Life Expectancy", y = "Frequency", 
       title = "Distribution of Life Expectancy by Consumption Level") +
  scale_fill_manual(values = c("High" = "darkblue", "Medium" = "blue", "Low" = "lightblue")) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, size = 11, face = "bold"))
```

### 3 Visualization

After gaining insights into the data distribution, we created visualizations to examine the association between life expectancy and alcohol consumption, and the influence of sex and time. A stacked histogram of alcohol consumption by sex showed that females have a smaller range and variance in alcohol consumption compared to males.

```{r, eval=TRUE, echo=FALSE, message=FALSE}
ggplot(df, aes(x = Consumption, fill = Sex)) +
  geom_histogram(bins = 30, alpha = 0.8, position = "stack", color = "white") +
  scale_fill_manual(values = c("pink", "lightblue")) + 
  labs(x = "Alcohol Consumption (liters per capita)", 
       y = "Frequency", 
       title = "Stacked Histogram of Alcohol Consumption by Sex") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, size = 11, face = "bold"))
```
Scatterplots with linear regression lines for 2000, 2010, and 2019 indicated a positive correlation between alcohol consumption and life expectancy—higher alcohol consumption was associated with longer predicted life expectancy. However, over time, the fitted line became less steep, suggesting that an increasing amount of alcohol consumption is needed to observe the same increase in life expectancy.

```{r, eval=TRUE, echo=FALSE, message=FALSE}
df_subset <- df |> filter(Year %in% c(2000, 2010, 2019))

ggplot(df_subset, aes(x = Consumption, y = `Life Expectancy`, color = as.factor(Year))) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE, color = "black") +
  facet_wrap(~ Year) +
  scale_color_manual(values = c("lightblue", "blue", "darkblue")) +
  labs(x = "Alcohol Consumption (liters per capita)", 
       y = "Life Expectancy (years)", 
       title = "Life Expectancy vs. Alcohol Consumption Over Decades") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, size = 11, face = "bold"),
        legend.position = "none")
```
We then conducted a regression analysis of life expectancy over time, adjusting for sex, using data from Canada and Japan. The results revealed a strong positive correlation between life expectancy and time, with an adjusted R² of approximately 0.99. Additionally, sex was a significant predictor, with Japan showing a larger life expectancy gap between females and males (-6.57 years) than Canada (-4.48 years).

```{r, eval=TRUE, echo=FALSE, message=FALSE}
df_subset <- df |> filter(Country %in% c("Canada", "Japan"))

ggplot(df_subset, aes(x = Year, y = `Life Expectancy`, color = Sex)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE) +
  facet_wrap(~ Country) +
  labs(x = "Year", y = "Life Expectancy", 
       title = "Life Expectancy Over Time: Canada vs. Japan") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, size = 11, face = "bold"),
        legend.position = "none")

df$year_centered <- df$Year - 2000
lm1 <- lm(`Life Expectancy` ~ year_centered + Sex, data = df[df$Country == "Canada", ])
lm2 <- lm(`Life Expectancy` ~ year_centered + Sex, data = df[df$Country == "Japan", ])
summary(lm1)
summary(lm2)
```
Next, we examined the top 10 countries with the largest life expectancy gaps in 2000 and 2019. The gap generally narrowed over time, except for the Central African Republic (CAF), which saw a drastic drop in both male and female life expectancy in 2019 (38.77 and 26.39 years, respectively).

```{r, eval=TRUE, echo=FALSE, message=FALSE}
df_gap_0 <- df |>
  filter(Year %in% c(2000)) |>
  group_by(Country) |>
  summarize(
    Life_Expectancy_f = sum(`Life Expectancy`[Sex == "Female"], na.rm = TRUE),
    Life_Expectancy_m = sum(`Life Expectancy`[Sex == "Male"], na.rm = TRUE)
  ) |>
  filter(Life_Expectancy_f > 0 & Life_Expectancy_m > 0) |>
  mutate(Gap = Life_Expectancy_f - Life_Expectancy_m) |>
  arrange(desc(Gap)) |>
  slice_head(n = 10)

df_gap_9 <- df |>
  filter(Year %in% c(2019)) |>
  group_by(Country) |>
  summarize(
    Life_Expectancy_f = sum(`Life Expectancy`[Sex == "Female"], na.rm = TRUE),
    Life_Expectancy_m = sum(`Life Expectancy`[Sex == "Male"], na.rm = TRUE)
  ) |>
  filter(Life_Expectancy_f > 0 & Life_Expectancy_m > 0) |>
  mutate(Gap = Life_Expectancy_f - Life_Expectancy_m) |>
  arrange(desc(Gap)) |>
  slice_head(n = 10)

df_gap <- bind_rows(
  df_gap_0 |> mutate(Year = 2000),
  df_gap_9 |> mutate(Year = 2019)
)

df_gap <- df_gap |>
  pivot_longer(cols = c(Life_Expectancy_f, Life_Expectancy_m, Gap), 
               names_to = "Sex", 
               values_to = "Life Expectancy")

df_gap <- df_gap |>
  mutate(Sex = factor(case_when(
    Sex == "Life_Expectancy_f" ~ "Female",
    Sex == "Life_Expectancy_m" ~ "Male",
    Sex == "Gap" ~ "Gap"
  ), levels = c("Female", "Male", "Gap")))

df_gap <- df_gap |>
  group_by(Year, Country) |>
  mutate(Gap_Value = `Life Expectancy`[Sex == "Gap"])

df_gap <- df_gap |> 
  mutate(Country = factor(Country, levels = unique(Country[order(-Gap_Value)])))
```

```{r, eval=TRUE, echo=FALSE, message=FALSE}
ggplot(df_gap, 
       aes(x = `Life Expectancy`, y = reorder(Country, Gap_Value), fill = Sex)) +
  geom_col(position = "dodge") +
  facet_wrap(~ Year, scales = "free_y") +
  geom_text(aes(label = round(`Life Expectancy`, 2)), 
          position = position_dodge(width = 0.8), 
          vjust = -0.5, size = 2.5) + 
  scale_fill_manual(values = c("Female" = "pink", "Male" = "lightblue", "Gap" = "gray")) +
  labs(y = "Country", x = "Life Expectancy (years)", 
       title = "Top 10 Countries with Largest Life Expectancy Gap by Sex") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, size = 11, face = "bold"))
```

```{r, eval=FALSE, echo=FALSE, message=FALSE}
### WRONG INTERPRETATION OF QUESTION 6 ###
df_gap_f <- df |>
  filter(Year %in% c(2000, 2019), Sex == "Female") |>
  group_by(Country) |>
  summarize(
    Life_Expectancy_2000 = sum(`Life Expectancy`[Year == 2000], na.rm = TRUE),
    Life_Expectancy_2019 = sum(`Life Expectancy`[Year == 2019], na.rm = TRUE)
  ) |>
  filter(Life_Expectancy_2000 > 0 & Life_Expectancy_2019 > 0) |>
  mutate(Gap = Life_Expectancy_2019 - Life_Expectancy_2000) |>
  arrange(desc(Gap)) |>
  slice_head(n = 10)

df_gap_m <- df |>
  filter(Year %in% c(2000, 2019), Sex == "Male") |>
  group_by(Country) |>
  summarize(
    Life_Expectancy_2000 = sum(`Life Expectancy`[Year == 2000], na.rm = TRUE),
    Life_Expectancy_2019 = sum(`Life Expectancy`[Year == 2019], na.rm = TRUE)
  ) |>
  filter(Life_Expectancy_2000 > 0 & Life_Expectancy_2019 > 0) |>
  mutate(Gap = Life_Expectancy_2019 - Life_Expectancy_2000) |>
  arrange(desc(Gap)) |>
  slice_head(n = 10)

df_gap <- bind_rows(
  df_gap_f |> mutate(Sex = "Female"),
  df_gap_m |> mutate(Sex = "Male")
)

ggplot(df_gap, aes(x = reorder(Country, Gap), y = Gap, fill = Sex)) +
  geom_col() +
  facet_wrap(~ Sex) +
  coord_flip() +
  labs(x = "Country", y = "Life Expectancy Discrepancy (years)", 
       title = "Top 10 Countries with Largest Life Expectancy Gap (2019 - 2000)") +
  scale_fill_manual(values = c("pink", "lightblue")) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, size = 11, face = "bold"),
        legend.position = "none")
```

A boxplot of life expectancy by alcohol consumption level and sex in 2019 showed that medium and low consumption levels resulted in similar life expectancies, while high alcohol consumption was associated with significantly longer life expectancy. Interestingly, females in the high-consumption group exhibited lower variance in life expectancy compared to other groups.

```{r, eval=TRUE, echo=FALSE, message=FALSE}
df_2019 <- df |> filter(Year == 2019)
ggplot(df_2019, aes(x = `Consumption Level`, y = `Life Expectancy`, fill = Sex)) +
  geom_boxplot() +
  scale_fill_manual(values = c("pink", "lightblue")) +
  labs(x = "Alcohol Consumption Level", 
       y = "Life Expectancy", 
       title = "Life Expectancy by Alcohol Consumption Level (2019)") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, size = 11, face = "bold"))
```

Finally, a scatterplot incorporating life expectancy, alcohol consumption, sex, and year illustrated that female data points were more clustered, reflecting their lower alcohol consumption variance. Additionally, life expectancy increased over time, as indicated by the progressive darkening of circle boundaries in the upper region of the data clusters. Among males, alcohol consumption appeared to decline over time, with darker-colored points concentrated on the left side of the plot.

```{r, eval=TRUE, echo=FALSE, message=FALSE}
ggplot(df, aes(y = `Life Expectancy`, x = Consumption, color = Year, fill = Sex)) +
  geom_point(alpha = 0.6, shape = 21) +
  facet_wrap(~ Sex) +
  scale_color_gradient(low = "pink", high = "purple") +
  scale_fill_manual(values = c("Female" = "orange", "Male" = "lightblue")) +
  labs(y = "Life Expectancy", x = "Alcohol Consumption", color = "Year", 
       title = "Life Expectancy vs. Alcohol Consumption Over Time") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, size = 11, face = "bold"))
```

### 4 Advanced Regression

To quantitatively support our findings, we applied multiple regression models. We first fit both linear and cubic models to examine the relationship between life expectancy and alcohol consumption, adjusting for scaled population, year, and sex. Both models yielded similar results. The adjusted R² values (0.243 for the linear model and 0.296 for the cubic model) confirmed our hypothesis that alcohol consumption explains only a small portion of life expectancy variation. The regression results showed that, on average, a female with low alcohol consumption in 2000 had a life expectancy of ~71 years (linear) or ~68 years (cubic), adjusted for population. Medium consumption levels had a minimal, statistically insignificant effect, while high consumption levels significantly increased life expectancy by ~7.97 years (linear) or ~7.61 years (cubic). Over time, females with low alcohol consumption experienced an increase in life expectancy (~0.3 years) per year. Males with low alcohol consumption had a statistically significantly lower life expectancy than females (-5.13 years) in the same year.

```{r, eval=TRUE, echo=FALSE, message=FALSE}
df <- df |>
  rename("Life_Expectancy" = "Life Expectancy",
         "Consumption_Level" = "Consumption Level")
```

```{r, eval=TRUE, echo=FALSE, message=FALSE}
# Linear Model
lm_mod <- lm(Life_Expectancy ~ Consumption_Level + log(Population) + year_centered + Sex, data = df)
summary(lm_mod)

# GAM
gam_mod <- gam(Life_Expectancy ~ Consumption_Level + s(log(Population), bs="cr") 
               + year_centered + Sex, data = df)
summary(gam_mod)
```

We also plotted the smooth term for the GAM model. We can see the first peak (around log(Population) ≈ 12) suggests that at lower population levels, life expectancy increases. The dip around log(Population) ≈ 16 could indicate a point where larger populations correlate with lower life expectancy, but this needs deeper investigation. The increase at the right end (high population levels) suggests that the largest countries tend to have higher life expectancy.

```{r, eval=TRUE, echo=FALSE, message=FALSE}
# Plot GAM
plot(gam_mod, se = TRUE, main = "Smooth term for Population")
```

We further examined the relationship by averaging alcohol consumption and life expectancy over time. The predictability of the models declined, with adjusted R² values dropping to 0.160 (linear) and 0.207 (cubic). The results showed that, on average, a female with zero alcohol consumption from 2000–2019 had a life expectancy of ~73 years (linear) or ~71 years (cubic), adjusted for population. A one-unit increase in average alcohol consumption raised female life expectancy by ~0.54 years (linear) or ~0.51 years (cubic). The male-female life expectancy gap increased to ~8.8 years (linear) or ~8.6 years (cubic).

```{r, eval=TRUE, echo=FALSE, message=FALSE}
# Remove Year
df_avg <- df |>
  group_by(Country, Sex) |>
  summarize(
    avg_consumption = mean(Consumption, na.rm = TRUE),
    avg_life_expectancy = mean(Life_Expectancy, na.rm = TRUE),
    avg_population = mean(Population, na.rm = TRUE)
  )

lm_avg <- lm(avg_life_expectancy ~ avg_consumption + log(avg_population) + Sex, data = df_avg)
summary(lm_avg)

gam_avg <- gam(avg_life_expectancy ~ avg_consumption + s(log(avg_population), bs="cr") 
               + Sex, data = df_avg)
summary(gam_avg)
```

These findings were reinforced by visualizations, where both linear and cubic regression lines indicated a positive relationship between average alcohol consumption and life expectancy. The cubic model better captured the relationship, particularly in the high-consumption region with a larger curve.

```{r, eval=TRUE, echo=FALSE, message=FALSE}
# Plot Both
ggplot(df_avg, aes(x = avg_consumption, y = avg_life_expectancy, color = Sex)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE, linetype = "dashed", color = "black") +
  geom_smooth(method = "gam", formula = y ~ s(x), se = TRUE, color = "purple") +
  labs(x = "Average Alcohol Consumption (liters per capita)", 
       y = "Average Life Expectancy (years)", 
       title = "Average Alcohol Consumption vs. Life Expectancy by Sex") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, size = 11, face = "bold"))
```

### Conclusion

Our analysis suggests a complex relationship between alcohol consumption and life expectancy. While higher alcohol consumption is associated with longer life expectancy, the effect size is small, and other factors likely play a more significant role. Additionally, the life expectancy gap between sexes has narrowed over time, though it remains substantial. Overall, our findings highlight the importance of considering multiple variables when analyzing life expectancy trends. Future research could explore additional socioeconomic and healthcare-related factors to better understand the drivers of life expectancy differences across countries and populations.