---
title: "Lab 01 - Hello R!"
author: "Sophie Xu"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: 
  html_document: 
    fig_width: 8
    fig_height: 4
    highlight: kate
    theme: cosmo
---

# Hello RStudio!

RStudio is comprised of four panes.

- On the bottom left is the Console, this is where you can write code that will be evaluated. Try typing `2 + 2` here and hit enter, what do you get?

- On the bottom right is the Files pane, as well as other panes that will come handy as we start our analysis.

- If you click on a file, it will open in the editor, on the top left pane.

- Finally, the top right pane shows your Environment. If you define a variable it would show up there. Try typing `x <- 2` in the Console and hit enter, what do you get in the Environment pane?

# Packages

R is an open-source language, and developers contribute functionality to R via packages. In this lab we will work with three packages: `datasauRus` which contains the dataset, and `tidyverse` which is a collection of packages for doing data analysis in a "tidy" way.

If you haven't installed these packages yet, you can install these packages by running the following command. (Note that R package names are case-sensitive)

```{r, message=FALSE, echo=FALSE}
install.packages(c("tidyverse", "datasauRus"))
```

Load them by running the following:

```{r message=FALSE, echo=FALSE}
library(tidyverse) 
library(datasauRus)
```


Note that the packages are also loaded with the same commands in your R Markdown document.

# Warm up

Before we introduce the data, let's warm up with some simple exercises.

The top portion of your R Markdown file (between the three dashed lines) is called YAML. It stands for "YAML Ain't Markup Language". It is a human friendly data serialization standard for all programming languages. All you need to know is that this area is called the YAML (we will refer to it as such) and that it contains meta information about your document.

## YAML

Open the R Markdown (Rmd) file in your project, change the author name to your name, and knit the document.

# Data

The data frame we will be working with today is called `datasaurus_dozen` and it's in the `datasauRus` package. Actually, this single data frame contains 12 datasets, designed to show us  why data visualisation is important and how summary statistics alone can be misleading. The different datasets are maked by the `dataset` variable.

To find out more about the dataset, type the following in your Console: `?datasaurus_dozen`. A question mark before the name of an object will always bring up its help file. This command must be ran in the Console.

1. Based on the help file, how many rows and how many columns does the `datasaurus_dozen` file have? What are the variables included in the data frame? Add your responses to your lab report. 

**Reponse: There are 1846 rows and 3 columns in the `datasaurus_dozen` dataset, with the variables `dataset`, `x`, and `y`.**

Let's take a look at what these datasets are. To do so we can make a *frequency table* of the dataset variable:

```{r, message=FALSE, eval=FALSE}
# Structure
dim(datasaurus_dozen)
glimpse(datasaurus_dozen)

# Twelve Datasets
datasaurus_dozen |>
  count(dataset)
```

The original Datasaurus (`dino`) was created by Alberto Cairo in [this great blog post](https://cran.r-project.org/web/packages/datasauRus/vignettes/Datasaurus.html). The other Dozen were generated using simulated annealing and the process is described in the paper *Same Stats, Different Graphs: Generating Datasets with Varied Appearance and Identical Statistics* through Simulated Annealing by Justin Matejka and George Fitzmaurice. In the paper, the authors simulate a variety of datasets that the same summary statistics to the Datasaurus but have very different distributions.

# Data visualization and summary

2. Plot `y` vs. `x` for the `dino` dataset. Then, calculate the correlation coefficient between `x` and `y` for this dataset.

Below is the code you will need to complete this exercise. Basically, the answer is already given, but you need to include relevant bits in your Rmd document and successfully knit it and view the results.

Start with the `datasaurus_dozen` and pipe it into the `filter` function to filter for observations where `dataset == "dino"`. Store the resulting filtered data frame as a new data frame called `dino_data`.

```{r}
dino_data <- datasaurus_dozen |>
  filter(dataset == "dino")
```

There is a lot going on here, so let's slow down and unpack it a bit. 

First, the pipe operator: `%>%`, which has been updated to `|>` (but both work), takes what comes before it and sends it as the first argument to what comes after it. So here, we're saying `filter` the `datasaurus_dozen` data frame for observations where `dataset == "dino"`.

Second, the assignment operator: `<-`, assigns the name `dino_data` to the filtered data frame.

Next, we need to visualize these data. We will use the `ggplot` function for this. Its first argument is the data you're visualizing. Next we define the `aes`thetic mappings. In other words, the columns of the data that get mapped to certain aesthetic features of the plot, e.g. the `x` axis will represent the variable called `x` and the `y` axis will represent the variable called `y`. Then, we add another layer to this plot where we define which `geom`etric shapes we want to use to represent each observation in the data. In this case we want these to be points,m hence `geom_point`.

```{r fig.fullwidth=TRUE}
ggplot(data = dino_data, mapping = aes(x = x, y = y)) +
  geom_point()
```

If this seems like a lot, it is. And you will learn about the philosophy of building data visualizations in layer in detail in the upcoming class. For now, follow along with the code that is provided.

For the second part of this exercises, we need to calculate a summary statistic: the correlation coefficient. Correlation coefficient, often referred to as $r$ in statistics, measures the linear association between two variables. You will see that some of the pairs of variables we plot do not have a linear relationship between them. This is exactly why we want to visualize first: visualize to assess the form of the relationship, and calculate $r$ only if relevant. In this case, calculating a correlation coefficient really doesn't make sense since the relationship between `x` and `y` is definitely not linear -- it's dinosaurial!

But, for illustrative purposes, let's calculate correlation coefficient between `x` and `y`.

Start with `dino_data` and calculate a summary statistic that we will call `r` as the `cor`relation between `x` and `y`.

```{r, message=FALSE}
dino_data |>
  summarize(r = cor(x, y))
```

3. Plot `y` vs. `x` for the `star` dataset. You can (and should) reuse code we introduced above, just replace the dataset name with the desired dataset. Then, calculate the correlation coefficient between `x` and `y` for this dataset. How does this value compare to the `r` of `dino`?

```{r, message=FALSE}
# Compute Correlation
star_data <- datasaurus_dozen |>
  filter(dataset == "star")

star_data |>
  summarize(r = cor(x, y))
```

**Reponse: The value r evaluates to -0.0629611, slightly less than the dino dataset's. Both of the correlation coefficients are relatively small.**


4. Plot `y` vs. `x` for the `circle` dataset. You can (and should) reuse code we introduced above, just replace the dataset name with the desired dataset. Then, calculate the correlation coefficient between `x` and `y` for this dataset. How does this value compare to the `r` of `dino`?

Facet by the dataset variable, placing the plots in a 3 column grid, and don't add a legend.

```{r, message=FALSE}
# Compute Correlation
circle_data <- datasaurus_dozen |>
  filter(dataset == "circle")

circle_data |>
  summarize(r = cor(x, y))
```

```{r}
# Three Plots
dino_data$dataset <- "Dino"
star_data$dataset <- "Star"
circle_data$dataset <- "Circle"
combined_data <- rbind(dino_data, star_data, circle_data)

# Create the faceted plot
ggplot(data = combined_data, mapping = aes(x = x, y = y)) +
  geom_point() +
  facet_wrap(~ dataset, ncol = 3) +
  theme(legend.position = "none")
```

**Reponse: The value r evaluates to -0.06834336, slightly larger than the dino dataset's. The correlation coefficient of the circle dataset is again relatively small.**


5. Finally, let's plot all datasets at once. In order to do this we will make use of facetting.

```{r all-viz, eval=FALSE, fig.fullwidth=TRUE}
ggplot(datasaurus_dozen, aes(x = x, y = y, color = dataset))+
  geom_point()+
  facet_wrap(~ dataset, ncol = 3) +
  theme(legend.position = "none")
```

And we can use the `group_by` function to generate all correlation coefficients.

```{r first-10-r, eval=FALSE}
datasaurus_dozen |>
  group_by(dataset) |>
  summarize(r = cor(x, y))
```

You're done with the data analysis exercises, but we'd like you to do two more things:

- **Resize your figures:** 

Click on the gear icon in on top of the R Markdown document, and select "Output Options..." in the dropdown menu. In the pop up dialogue box go to the Figures tab and change the height and width of the figures, and hit OK when done. Then, knit your document and see how you like the new sizes. Change and knit again and again until you're happy with the figure sizes. Note that these values get saved in the YAML.

You can also use different figure sizes for different figures. To do so click on the gear icon within the chunk where you want to make a change. Changing the figure sizes added new options to these chunks: `fig.width` and `fig.height`. You can change them by defining different values directly in your R Markdown document as well.

- **Change the look of your report:** 

Once again click on the gear icon in on top of the R Markdown document, and select "Output Options..." in the drop-down menu. In the General tab of the pop up dialogue box try out different syntax highlighting and theme options. Hit OK and knit your document to see how it looks. Play around with these until you're happy with the look.


# Optional

If you have time you can explore the different ways you can add styling to your rmarkdown document.

Here is a [Rmarkdown cheatsheet](https://rstudio.github.io/cheatsheets/html/rmarkdown.html)

and a [general markdown cheatsheet](https://www.markdownguide.org/cheat-sheet/) and some [themes](https://rpubs.com/ranydc/rmarkdown_themes) and [more themes](https://bookdown.org/yihui/rmarkdown/appearance-and-style-1.html)

---

This set of lab exercises have been adopted from [Mine Çetinkaya-Rundel](https://mine-cr.com/)'s class [Introduction to Data Science](https://sta199-s24.github.io/).

