---
title: "Lab 9 - HPC"
output: html_document
link-citations: yes
---

# Learning goals

In this lab, you are expected to practice the following skills:

- Evaluate whether a problem can be parallelized or not.
- Practice with the parallel package.
- Use Rscript to submit jobs.

```{r eval=TRUE, echo=FALSE}
# install any missing packages
#install.packages("microbenchmark")
library(parallel)
```

## Problem 1

Give yourself a few minutes to think about what you learned about parallelization. List three
examples of problems that you believe may be solved using parallel computing,
and check for packages on the HPC CRAN task view that may be related to it.

* `doParallel`, `mlr`, `foreach`, `foParallel` -> supports cross-validation in machine learning and helps parallelize model training 
* `boot`, `parallel` -> supports bootstrapping, and parallelize re-sampling
* `parallel`, `RcppParallel`, `rstan`, `nimle` -> assists in implementing Markov Chain Monte Carlo, and Bayesian modeling and inferences

## Problem 2: Pre-parallelization

The following functions can be written to be more efficient without using
`parallel`:

1. This function generates a `n x k` dataset with all its entries having a Poisson distribution with mean `lambda`.

```{r p2-fun1, eval = TRUE}
fun1 <- function(n = 100, k = 4, lambda = 4) {
  x <- NULL
  
  for (i in 1:n)
    x <- rbind(x, rpois(k, lambda))
  
  return(x)
}

fun1alt <- function(n = 100, k = 4, lambda = 4) {
  # pre-allocate memory
  matrix(rpois(n * k, lambda = lambda), ncol = k)
}

# Benchmarking
microbenchmark::microbenchmark(
  fun1(),
  fun1alt(),
  unit = "ns"
)
```

How much faster?

_The second function utilizing matrices is 10x faster than the first one._


2.  Find the column max (hint: Checkout the function `max.col()`).

```{r p2-fun2, eval = TRUE}
# Data Generating Process (10 x 10,000 matrix)
set.seed(1234)
x <- matrix(rnorm(1e4), nrow=10)

# Find each column's max value
fun2 <- function(x) {
  apply(x, 2, max)
}

fun2alt <- function(x) {
  # YOUR CODE HERE
  x[cbind(max.col(t(x)), 1:ncol(x))]
}

# Benchmarking
bench <- 
  microbenchmark::microbenchmark(
    fun2(x),
    fun2alt(x),
    unit = "us"
  )
```

```{r}
plot(bench)
ggplot2::autoplot(bench) +
ggplot2::theme_minimal()
```

_We can see the second function utilizing matrices is again significantly faster than the first one._


## Problem 3: Parallelize everything

We will now turn our attention to non-parametric 
[bootstrapping](https://en.wikipedia.org/wiki/Bootstrapping_(statistics)).
Among its many uses, non-parametric bootstrapping allow us to obtain confidence
intervals for parameter estimates without relying on parametric assumptions.

The main assumption is that we can approximate many experiments by resampling
observations from our original dataset, which reflects the population. 

This function implements the non-parametric bootstrap:

```{r p3-boot-fun, eval = TRUE}
my_boot <- function(dat, stat, R, ncpus = 1L) {
  
  # Getting the random indices
  n <- nrow(dat)
  idx <- matrix(sample.int(n, n*R, TRUE), nrow=n, ncol=R)
 
  # Making the cluster using `ncpus`
  # STEP 1: GOES HERE
  cl <- makePSOCKcluster(ncpus)
  
  # STEP 2: GOES HERE
  # send variables to worker nodes 
    # note they are running in isolated environments, so there are no global variables
    # idx -> re-sampling indices, dat -> dataset, stat -> function to compute estimates
  clusterExport(cl, varlist = c("idx", "dat", "stat"), envir = environment()) 
  
  # STEP 3: THIS FUNCTION NEEDS TO BE REPLACED WITH parLapply
  ans <- parLapply(cl, seq_len(R), function(i) {
    stat(dat[idx[,i], , drop=FALSE])
  })
  
  # Coercing the list into a matrix
  ans <- do.call(rbind, ans)
  
  # STEP 4: GOES HERE
  stopCluster(cl) # free up system resources
  # on.exit(stopCluster(cl)) # prevents memory leak
  ans
  
}
```

1. Use the previous pseudocode, and make it work with `parallel`. Here is just an example for you to try:

```{r p3-test-boot, eval = TRUE}
# Bootstrap of a linear regression model
my_stat <- function(d) coef(lm(y ~ x, data = d))

# DATA SIM
set.seed(1)
n <- 500 
R <- 1e4
x <- cbind(rnorm(n)) 
y <- x*5 + rnorm(n)

# Check if we get something similar as lm
ans0 <- confint(lm(y ~ x))
cat("OLS CI \n")
print(ans0)

ans1 <- my_boot(dat = data.frame(x, y), my_stat, R = R, ncpus = 4)
qs <- c(.025, .975)
cat("Bootstrap CI \n")
print(t(apply(ans1, 2, quantile, probs = qs)))

```

2. Check whether your version actually goes faster than the non-parallel version:

```{r benchmark-problem3, eval = TRUE}
# your code here
detectCores()

# non-parallel
system.time(my_boot(dat = data.frame(x, y), my_stat, R = 4000, ncpus = 1L))

# parallel 8 cores
system.time(my_boot(dat = data.frame(x, y), my_stat, R = 4000, ncpus = 8L))
```

_The parallel version indeed runs faster than the non-parallel version._

## Problem 4: Compile this markdown document using Rscript

Once you have saved this Rmd file, try running the following command
in your terminal:

```bash
Rscript --vanilla -e 'rmarkdown::render("C:\Users\xsoph\OneDrive\Documents\RStudio Projects\JSC370-Coursework\lab09\lab09-hpc.Rmd")' &
```

Where `[full-path-to-your-Rmd-file.Rmd]` should be replace with the full path to
your Rmd file... :).


